---
title: "逆高斯退化过程"
author: "庄亮亮"
format: 
  html:
    toc: TRUE
    toc-depth: 5
    toc-title: "目录"
    theme: cerulean
    code-tools: true
    highlight-style: cerulean.theme
    code-line-numbers: true
    code-block-bg: true
editor: visual
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## 1. IG 退化过程模型

虽然 Wiener 过程和 Gamma 过程在退化建模中得到了广泛的应用，但这两种模型在实际应用中无法拟合大量退化数据。当这两个过程失败时，另一个选择是逆高斯(inverse Gaussian, IG)过程。

过程$\{Y(t);t \geqslant 0\}$如果具有独立的IG分布增量，则称为基本IG过程，即$Y(t) \sim \operatorname{IG}\left(\mu \Lambda(t)， \Lambda \Lambda^2(t)\right)$，其中$\Lambda(t)$是非负单调递增的。IG $(a, b), a, b>0$ 的PDF为 $$
f_{\mathrm{IG}}(y ; a, b)=\left(\frac{b}{2 \pi y^3}\right)^{1 / 2} \exp \left[-\frac{b(y-a)^2}{2 a^2 y}\right], y>0
$$ 在Wang和Xu\[48\]最近的一项研究之前，IG过程很少被用作退化模型。这可能是因为这个过程没有物理解释。因此，可靠性工程师可能不知道如何以及何时使用该过程进行退化建模。为了弥补这一差距，Ye and Chen \[49\]探索了这一过程的物理意义。他们发现IG过程与Gamma过程类似，也是复合泊松过程的极限，但震动大小分布与Gamma过程不同。此外，他们还发现，与Gamma过程相比，IG过程在合并协变量和随机效应方面非常灵活。灵活性来自于Wiener过程和IG过程之间的逆关系，如下所示。

回想一下，维纳过程$X(u)=v u+\sigma \mathrm{B}(u)$到一个固定阈值的第一次传递时间$\Lambda$是一个IG随机变量$Y \sim \operatorname{IG}\left(\Lambda / v， \Lambda^2 / \sigma^2\right)$。如果我们有一系列由$t$索引的阈值$\Lambda$，我们将有一系列对应的失败次数$Y(t)$。很容易表明$Y(t)$遵循一个IG过程$Y(t) \sim \operatorname{IG}\left(\mu \Lambda(t)， \Lambda \Lambda^2(t)\right)$，其中$\mu=1 / \nu$和$\ Lambda =1 / \sigma^2$。有趣的是，只要Wiener过程中有一种方法可以合并协变量和随机效应，那么IG过程中也有相应的方法可以通过逆关系来合并协变量和随机效应。例如，在维纳过程中，我们可以假设$\nu$是正常的，$\sigma$是固定的，或者我们可以假设$\sigma^{-2} \sim \operatorname{Gamma}(r， \delta)$和$\left[v \mid \sigma^2\right] \sim \mathrm{N}\left(v， \theta \sigma^2\right)$。因此，我们可以假设$\mu=1 / v$是正常的，或者假设在IG进程中$\lambda=\sigma^{-2} \sim \operatorname{Gamma}(r,\delta)$和$[1 / \mu \mid \lambda] \sim \mathrm{N}(v， \theta / \lambda)$。由此得到的IG随机效应模型的边际分布是可处理的。类似地，在$2.3$中回顾的Wiener过程中合并协变量的方法可以通过逆关系扩展到IG过程。

在Wang和Xu\[48\]的工作中，针对IG过程提出了一种基于EM算法的半参数估计方法，其中函数$\Lambda(t)$非参数估计。EM算法需要IG过程的条件增量的期望，这是很难得到的。在Ye\[50\]中导出了期望的闭合表达式。这些结果为IG过程的EM算法的实现提供了便利。Peng等人\[51\]为IG过程开发了贝叶斯方法。然而，IG过程在退化文献中仍然是新的。退化模拟过程的性质需要进一步探索。

### 3.1 经典的 IG 过程

> 来自：Guan, Q., et al. (2019). "Reference Bayesian analysis of inverse Gaussian degradation process." Applied Mathematical Modelling 74: 496-511.（IG + 客观贝叶斯）

Let $Y(t)$ be the degradation characteristic of the product at time $t$. Suppose that the degradation path $Y(t)$ can be modeled by an IG process with mean function $\Lambda(t)$ and the scale parameter $\lambda$. Thus, $Y(t)$ have the following properties: - $Y(0)=0$ with probability one; - For $t>s>u, Y(t)-Y(s) \geq 0, Y(s)-Y(u) \geq 0, Y(t)-Y(s)$ and $Y(s)-Y(u)$ are independent; - For $t>s \geq 0$, each increment $Y(t)-Y(s)$ has an IG distribution $I G\left(\Lambda(t)-\Lambda(s), \lambda(\Lambda(t)-\Lambda(s))^2\right)$, where $\Lambda(t)$ is a monotonic increasing function, and $\Lambda(0)=0$. The probability density function (pdf) of an IG distribution $I G(a, b)$ is $$
f_{I G}(x, a, b)=\left[\frac{b}{2 \pi x^3}\right]^{1 / 2} \exp \left\{-\frac{b(x-a)^2}{2 a^2 x}\right\}
$$ When $\Lambda(t)$ is a linear function, $\Lambda(t+\Delta t)-\Lambda(t)=\Lambda(s+\Delta t)-\Lambda(s)$ holds for any $s \geq 0, t \geq 0$, and $\Delta t \geq 0$. That is, this IG process is a stationary process. When $\Lambda(t)$ is a non-linear function, $\Lambda(t+\Delta t)-\Lambda(t)$ depends on $t$ and $\Delta t$. The corresponding IG process is a non-stationary process. In this study, we suppose that $\Lambda(t)=\eta t$. It is easy to compute the expectation and variance of $Y(t+\Delta t)-Y(t)$ according to the property of IG process, that is, $$
E[Y(t+\Delta t)-Y(t)]=\Lambda(t+\Delta t)-\Lambda(t)=\eta \Delta t,
$$ $$
\operatorname{Var}[Y(t+\Delta t)-Y(t)]=\frac{\Lambda(t+\Delta t)-\Lambda(t)}{\lambda}=\frac{\eta \Delta t}{\lambda} .
$$ The lifetime $T$ can be defined as the first passage time when $Y(t)$ crosses a critical value $\omega$. Hence, we have $$
T=\inf \{t \mid Y(t) \geq \omega, t>0\} .
$$ For given $\omega$, the lifetime $T$ has the following cumulative distribution function (cdf) $$
\begin{aligned}
F(t) &=P(T \leq t)=P(Y(t)>\omega)=1-F_{I G}\left(\omega ; \Lambda(t), \lambda \Lambda(t)^2\right) \\
&=\Phi\left[\sqrt{\frac{\lambda}{\omega}}(\Lambda(t)-\omega)\right]-e^{2 \lambda \Lambda(t)} \Phi\left[-\sqrt{\frac{\lambda}{\omega}}(\Lambda(t)+\omega)\right]
\end{aligned}
$$ where $F_{I G}(. ; a, b)$ is cdf of the IG distribution, and $\Phi(\cdot)$ is cdf of the standard normal distribution.

Taking the derivative of $(3)$ with respect to $t$, the lifetime $T$ has the following pdf $$
\begin{aligned}
f(t)=& \Lambda^{\prime}(t)\left\{\frac{\lambda}{\omega} \phi\left[\frac{\sqrt{\lambda}(\Lambda(t)-\omega)}{\sqrt{\omega}}\right]-2 \lambda \exp [2 \lambda \Lambda(t)] \Phi\left[-\sqrt{\frac{\lambda}{\omega}}(\Lambda(t)+\omega)\right]\right\} \\
&+\Lambda^{\prime}(t)\left\{\frac{\lambda}{\omega} \exp [2 \lambda \Lambda(t)] \phi\left[-\sqrt{\frac{\lambda}{\omega}}(\Lambda(t)+\omega)\right]\right\}
\end{aligned}
$$ where $\phi(\cdot)$ is pdf of standard normal distribution. If $t \rightarrow \infty$, then $\lambda \Lambda(t) \rightarrow \infty$ and $Y(t)$ follows an asymptotic normal distribution with mean $\Lambda(t)$, and variance $\frac{\Lambda(t)}{\lambda}$. Therefore, the cdf of lifetime $T$ is asymptotically $$
P(T<t) \approx 1-\Phi\left[\frac{\omega-\Lambda(t)}{\sqrt{\Lambda(t) / \lambda}}\right]=\Phi\left[\sqrt{\lambda \Lambda(t)}-\frac{\omega \sqrt{\lambda}}{\sqrt{\Lambda(t)}}\right]
$$ When $\Lambda(t)=\eta t$ \[17\], Eq. (5) is BS distribution with shape parameter $(\omega \lambda)^{-\frac{1}{2}}$ and scale parameter $\frac{\omega}{\eta}$. That is, the cdf of lifetime $T$ is an asymptotic BS distribution. Thus, the mean time to failure (MTTF) of the lifetime $T$ is the mean of BS distribution, that is, $$
M T T F=E(T) \approx \frac{\omega}{\eta}+\frac{1}{2 \eta \lambda} .
$$ BS distribution as an effective lifetime distribution for fatigue failure time data has been widely studied \[18-22\]. However, BS distribution is not proper to be used as a degradation model and could not replace the IG process model.

```{r}
library(ggplot2)
library(tidyverse)
library(viridis)
```

#### 3.1.1 数据模拟产生

```{r}
# 数据模拟产生
sim_dat = function(group = 6, t = 1:200, para = c(1,0.1)){
  # para 指 mu,sigma
  # epoch 指测量次数， group 指组数
  delta_t = diff(t)
  epoch = length(delta_t)
  dat = matrix(NA,epoch,group+1)
  dat[,1] = seq(1,epoch)
  dat_unit = numeric()
  for(i in 1:group){
    for(j in 1:epoch) dat_unit[j] = rnorm(1,para[1]*delta_t[j],sqrt(para[2]^2*delta_t[j]))
    dat[,i+1] = cumsum(dat_unit)
  }
  dat = data.frame(dat)
  dat1 = rbind(rep(0,group+1),dat) # 加入初始值绘制点
  colnames(dat1) = c("Time",paste(1:group,sep=''))
  return(dat1)
}
```

绘制退化路径图

```{r}
# 绘制退化路径图
plot_path = function(data = dat1){
    # 画图
  p = data %>% pivot_longer(paste(1:group), 
                       names_to = "Group", 
                       values_to = "y") %>% 
    ggplot(aes(Time,y,color = Group)) + 
    geom_line() + 
    scale_color_viridis(discrete = T) + 
    ylab("Degradation") #+
    #theme_bw() +
    #theme(panel.grid = element_blank())
  print(p)
}
```

```{r}
group = 6
t = 1:200
para = c(1,1)
dat = sim_dat(group = group, t = t, para = para)
knitr::kable(head(dat))
plot_path(dat)
```

#### 3.1.2 统计推断

We define $Y\left(t_{i j}\right)$ to be the degradation characteristic measured at time $t_{i j}$ for the $i$ th unit, where $i=1, \ldots, n$ and $j=$ $1, \ldots, m_i$. According to $(1)$, based on $\Delta y_{i j}=Y\left(t_{i j}\right)-Y\left(t_{i j-1}\right), i=1, \ldots n, j=1, \ldots, m_i$, the likelihood function of $(\lambda, \eta)$ is $$
L(\lambda, \eta)=\prod_{i=1}^n \prod_{j=1}^{m_i} f_{I G}\left(\Delta y_{i j}, \eta \Delta t_{i j}, \lambda \eta^2 \Delta^2 t_{i j}\right) .
$$

##### 3.1.2.1 极大似然估计

Lemma 1. The Fisher information matrix of $(\eta, \lambda)$ is $$
I(\eta, \lambda)=\left[\begin{array}{cc}
\frac{\lambda \eta \tau+2 N}{\eta^2} & \frac{N}{\lambda \eta} \\
\frac{N}{\lambda \eta} & \frac{N}{2 \lambda^2}
\end{array}\right]=\left[\begin{array}{cc}
I_{11}(\eta, \lambda) & I_{12}(\eta, \lambda) \\
I_{21}(\eta, \lambda) & I_{22}(\eta, \lambda)
\end{array}\right],
$$ where $\tau=\sum_{i=1}^n \sum_{j=1}^{m_i} \Delta t_{i j}$ and $N=\sum_{i=1}^n m_i$. Proof. The loglikelihood function is $$
\log L(\lambda, \eta)=\sum_{i=1}^n \sum_{j=1}^{m_i}\left(\frac{1}{2} \log (\lambda)+\log (\eta)+\log \Delta t_{i j}-\frac{1}{2} \log \left(2 \pi \Delta y_{i j}^3\right)-\frac{\lambda\left(\Delta y_{i j}-\eta \Delta t_{i j}\right)^2}{2 \Delta t_{i j}}\right)
$$

```{r}
cumsub = function(x){
  #完成逐个相减功能，类似cumsum
  re = numeric()
  re[1] = x[1]
  for(i in 2:length(x)){
    re[i] = x[i] - x[i-1]
  }
  return(re)
}
```

```{r}
Weiner_mle = function(par = c(1,1),data = dat){
  # 输入：mu，sigma，数据
  # 输出：负对数似然函数
  mu = par[1]; sigma = par[2]
  group = ncol(data) - 1;time = data[,1];y = data[-1,-1]
  # 差分得到delta_time和delta_y
  delta_time = matrix(rep(diff(time),group),length(diff(time)),group)
  delta_y = matrix(NA,nrow(delta_time),group)
  for(i in 1:group) delta_y[,i] = cumsub(y[,i])

  logl = -1/2 * sum(log(delta_time)) - group*nrow(delta_time)*log(sigma) - sum((delta_y - mu * delta_time)^2/(2*sigma^2*delta_time),na.rm =T)

  return(-logl)
}

mle_re = optim(par=c(1,1), fn = Weiner_mle, 
      data = dat, method = "BFGS", hessian = TRUE)
# mle_re
mle_par = mle_re$par
# 区间估计
mle_up = mle_re$par + sqrt(diag(solve(mle_re$hessian)))*qnorm(0.975)
mle_low = mle_re$par - sqrt(diag(solve(mle_re$hessian)))*qnorm(0.975) #hessian 和协方差矩阵存在逆的关系：https://blog.csdn.net/anruoxi3236/article/details/102005217
mle_summary = round(cbind(mle_low,mle_re$par,mle_up),3)
colnames(mle_summary) = c("low","mean","up")
knitr::kable(mle_summary) 
```

##### 3.1.2.2 客观贝叶斯分析

> 来自：Guan, Q., et al. (2019). "Reference Bayesian analysis of inverse Gaussian degradation process." Applied Mathematical Modelling 74: 496-511.（IG + 客观贝叶斯）

##### 3.1.2.3 广义推断方法

暂时还未找到，最经典的维纳过程的广义推断方法论文。其他类似的论文整理如下：

### 3.3 案例分析

#### Laser device example

#### Anti-radiation example
